Comparison of particle trajectories and collision operators for collisional transport in nonaxisymmetric plasmas

### 2014-02-21 doc/sfincsPaper/refereeResponse/sfincsPaper_refereeResponse.tex

We have added a new section IV to the paper to discuss momentum balance.
To summarize, we first discuss how parallel momentum has a different status to mass and energy in that mass and
energy are conserved, whereas parallel momentum is not conserved due to the mirror force.
We then compare the $m v_{||}$ moments of the various forms of the drift-kinetic
equation to the full fluid momentum equation, as you suggest.
We show the electric field terms in the kinetic perspective
are associated with a piece of the gyroviscosity in the fluid perspective.  The full trajectory model
gives the correct viscous force whereas the other trajectory models do not.
One further noteworthy property of the parallel velocity moments of the kinetic equations
is that of the trajectory models considered, only the full trajectories preserve intrinsic
ambipolarity in quasisymmetry when $E_r \ne 0$. This result is now discussed at the end of section IV.

\begin{referee}
I do not understand why the source term (25) does not work well with the pitch-angle scattering
operator or P.-A. scattering + momentum conservation operator. It is not the difference in the
collision operator but difference in the trajectory model that is related to the conservation
property of energy, (22). If the original motivation of introducing the term is to maintain the
mass and energy conservation property for finite-Er, then why you need to change the form of
source/sink term if you choose P.-A. scattering operator, which actually conserves the energy,
while you do not need according to the trajectory model? Does it mean diffusion process in
v-coordinate is essential for the source/sink term to work properly?
Please explain the reason of changing the source term according to the collision operator more
detail.
\end{referee}

We have expanded the discussion at the end of section III to clarify this issue.

The system of sources and extra constraint equations actually serves two independent purposes,
not only to circumvent the conservation problems when $E_r$ is nonzero, but also to eliminate the null
space of the kinetic equation. This latter problem arises even for the DKES trajectories 
and even at $E_r=0$.  
When $E_r \ne 0$, the sources are needed to eliminate the conservation problems, and the extra constraints 
can be thought of as a convenient way to keep the linear system square (number of equations = number of unknowns) upon discretizatiton.
Even when $E_r = 0$, and even for the DKES trajectories in which sources are not required, the constraints are needed to eliminate
the null space in the kinetic equation, and the source terms can be thought of as
a convenient way to keep the linear system square upon discretization.
Any particle and heat source will be sufficient to solve the first problem (conservation), but to solve the second problem (null space),
the number of constraints should match the dimensionality of the null space.  This dimension is set by the collision operator (2 for
Fokker-Planck, $N_x$ for the other two operators.)
This is why the source/constraint scheme is chosen based on the collision operator.

### 2014-02-21 doc/sfincsPaper/sfincsPaper.tex
Finished draft of referee response and manuscript corrections

@@

The aforemention mass and energy sources and constraints work when the Fokker-Planck collision operator
is used, 
but not when the pure pitch-angle scattering or momentum-conserving model collision operator is used, since in these cases the kinetic equation has a larger null space
(any function of $v$ is a solution.)
For these collision operators, we instead allow a general source $S_a(\psi,v)$,
introducing $N_x$ unknowns on each flux surface, where $N_x$ is the number of grid points in $v$.
To supply an equal number of constraints, we impose the equations $\left< \int_{-1}^1 d\xi\, f_{a1}\right>=0$
at each grid point in $v$.

>>>>

\changed{
Note that for both the DKES and full trajectories, the left-hand side of the
kinetic equation (\ref{eq:kinetic4})
can be written in the conservative form
\begin{equation}
\frac{1}{J}\left[
\nabla\cdot\left( J \dot{\vect{r}}_a f_{a1} \right) 
+ \frac{\partial}{\partial \xi}\left( J \dot{\xi}_a f_{a1} \right)
+ \frac{\partial}{\partial x_a} \left( J \dot{x}_a f_{a1} \right)
\right] - C_a
\label{eq:conservativeForm}
\end{equation}
where $J=x_a^2$ is the Jacobian of the transformation between Cartesian velocity coordinates
and the coordinates $x_a$, $\xi$, and gyrophase. However, for the partial trajectories,
the left-hand side of (\ref{eq:kinetic4}) is not equivalent
to (\ref{eq:conservativeForm}).
}

@@

\changed{
This system of sources and constraints solves not only the problem described above when $E_r \ne 0$,
but also a different problem that remains even when $E_r=0$ and/or when the DKES trajectories are used:
the kinetic equation has a null space.
If the conditions $\left< \int d^3v \,f_{a1}\right>=0$ and $\left< \int d^3v\, v^2 f_{a1}\right>=0$
were not imposed, any linear combination of $F_{a}$ and $F_{a} v^2$ could be added to one solution
of the kinetic equation to obtain another solution. Upon discretization, one would obtain a non-invertible
(or at least very poorly conditioned) linear system, but the imposition of these two extra constraints
makes the system of equations invertible. 
}

\changed{
Such is the case when the full linearized Fokker-Planck 
collision operator is used, but the situation is different when either the pitch-angle scattering
operator or momentum-conserving model operator are used instead, for then the kinetic equation has a larger null space:
any function of $v$ is then a homogeneous solution of the kinetic equation.
As the dimension of the null space is then equal to $N_x$ (the number of grid points in $x_a$) rather than
2, it takes $N_x$ rather than 2 constraint equations to eliminate the null space for these collision operators.
We choose these $N_x$ constraints to be $\left< \int_{-1}^1 d\xi\, f_{a1}\right>=0$ (imposed at each grid point in $x_a$.)
To keep the linear system square, we must then have $N_x$ rather than 2 unknowns related to the sources. 
This is accomplished by letting the source be a general function of $x_a$ instead
of (\ref{eq:sources}) when either the pitch-angle scattering or momentum-conserving model collision operator are used.
This alternative system of $N_x$ sources and constraints is an equally reasonable solution to the earlier
conservation problem.
}

\changed{
To summarize, the sources and extra constraint equations serve two independent purposes.
First, when $E_r \ne 0$, the sources are needed to eliminate the conservation problems, and the extra constraints then keep the linear system square (number of equations = number of unknowns) upon discretizatiton.
Second, even when $E_r = 0$, and even for the DKES trajectories in which sources are not required, the constraints are needed to eliminate the null space in the kinetic equation, and the source terms are a convenient way to keep the linear system square upon discretization.
The first problem can be solved with either the source (\ref{eq:sources}) or a general speed-dependent source $S_a(x_a)$.
However, to solve the second problem, the number of constraints should match the dimensionality of the null space.
For this reason, we apply the source (\ref{eq:sources}) with 2 constraints when the Fokker-Planck operator is used,
while we apply the general speed-dependent source $S_a(x_a)$ with $N_x$ constraints when either of the other two collision operators is used.
}

\begin{comment}
The aforemention mass and energy sources and constraints work when the Fokker-Planck collision operator
is used, 
but not when the pure pitch-angle scattering or momentum-conserving model collision operator is used, since in these cases the kinetic equation has a larger null space
(any function of $v$ is a solution.)
For these collision operators, we instead allow a general source $S_a(\psi,v)$,
introducing $N_x$ unknowns on each flux surface, where $N_x$ is the number of grid points in $v$.
To supply an equal number of constraints, we impose the equations $\left< \int_{-1}^1 d\xi\, f_{a1}\right>=0$
at each grid point in $v$.
\end{comment}
